Density-dependent selection and the limits of relative fitness

### 2018-04-13 manuscript.tex
l

Another issue with the constant-$N$, constant-$s$ description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. A variety of approaches have been developed to address this in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}. Although density-dependent selection is  pertinent to this longer-term issue, our focus here is the description of the time-dependent process by which selection changes allele frequencies. This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.

@@

In light of the complications of density-dependence, the assignment of density-independent relative fitnesses has been justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}). Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig. 9.5]{otto_2011}.

@@

Here we argue that the simplified models of density-regulated growth mentioned above are misleading in their representation of the interplay between selection and density. This ultimately derives from their failure to account for ``reproductive excess'', that is, an excess of juveniles that experience stronger selection than their adult counterparts \citep{turner1968population}. This allows selection to be concentrated at a juvenile ``bottleneck'', potentially keeping the density of adults constant even under strong selection. Reproductive excess featured prominently in early debates about the regulation of population density (e.g. \citealt{nicholson_1954}), and also has a long history in evolutionary theory, particularly related to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}. Additionally, reproductive excess is implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each generation involves the production of an infinite number of zygotes, of which a constant number $N$ are sampled to form the next generation of adults. Likewise in the Moran model, a juvenile is always available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.

@@

In models with detailed age structure, it is usually assumed that the density of a ``critical age group'' mediates the population's response to crowding \citep[pp. 54]{charlesworth_1994}. Reproductive excess is a special case corresponding to a critical pre-reproductive age group. A central result of the theory of density-regulated age-structured populations is that selection proceeds in the direction of increasing equilibrium density in the critical age group \citep[pp. 148]{charlesworth_1994}. This is a form of the classical $K$-selection ideas discussed above, but restricted to the critical age group (juveniles, in this case). By focusing on density in the critical age group, this literature loses the interdependence of pre-reproductive selection and post-reproductive density.

We develop a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a finite reproductive excess. Our starting point is the classic lottery model, which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. Each type is assumed to have three traits: fecundity $b$, mortality $d$, and competitive ability $c$. In each iteration of the classic lottery model, each type produces a large number of juveniles, such that $N$ remains constant (infinite reproductive excess). Competitive ability $c$ affects the probability of winning a territory, and behaves like a pure relative fitness trait. Thus, fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). Our first task is to relax the large-juvenile-number assumption of the lottery model to create a variable-density lottery with a finite, density-dependent reproductive excess.

The properties of density-dependent selection in our model are strikingly different from the classical literature discussed above. The strong connection between crowding and selection for greater equilibrium density is broken: selection need not affect density, and even when it does, Eq.~\eqref{eq:canonical} is almost exact even for strong selection, provided that only one of the traits $b$, $c$, or $d$ is involved.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the difference in propagule success rate $\Delta n_2/m_2-\Delta n_1/n_1$ between an adaptive variant $2$ and a wildtype $1$ with equal frequencies. Here $b_1=1$, $d_1=0.5$ and $c_1=1$. For $b$-selection we set $b_2=b_1(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as  available territories become scarce. The wildtype equilibrium density is $n_1/T\approx 0.4$.}
\end{figure}

@@

Compared to simple birth-death models \citep{kostitzin_1939} or variants of the logistic \citep{roughgarden_1979}, the Lotka-Volterra model clearly distinguishes between intra- and inter-type competitive effects. Thus, one type can displace another without having a greater equilibrium density (Fig.~\ref{fig:LVvslottery}a). Nevertheless, selection drives transient changes in density in the Lotka-Volterra model even if the initial and final densities of a sweep are the same (constant density only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012,smouse_1976}). Intuitively, for one type to exclude the other, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing $N$ to dip over a sweep (Fig.~\ref{fig:LVvslottery}a).

@@

Summarizing the three traits in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet despite their differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the density-independent relative fitness description of selection almost exactly. This behavior is quite different from that found in the classical density-dependent selection literature (see ``Introduction'' and ``The response of density to selection; $c$-selection versus $K$-selection''). In the latter, selection in crowded populations takes broadly one of two forms: selection for greater carrying capacity ($K$-selection) or selection on competition coefficients (``$\alpha$-selection''; \citealt{gill_1974,joshi_2001}). $K$ and $\alpha$ both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that strong selection is sufficient for Eq.~\eqref{eq:canonical} to break down (Fig.~\ref{fig:strengthofselection}).

In our variable-density lottery, $\delta$-like behavior occurs if and only if types differ in more than one of the three traits $b$, $c$ and $d$ (Fig.~\ref{fig:multiple}). The $c$ and $d$ traits embody the two distinct directions in which density and selection can interact: selection may depend on density, and density may depend on selection \citep{prout_1980}. The combination of both is necessary to threaten the constant-$s$ approximation. The simple differential/difference equations that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} confound this distinction. Remarkably, the $b$ trait demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.

While this is a conceptual reason to be wary of the traits analyzed in classical density-dependent selection models, it is not clear what trait variation should be expected in nature. Should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent? In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent. However, for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally. In this case, both resource competition and interference competition are subsumed into the competitive ability $c$, which, in the absence of pleiotropic effects on $b$ and/or $d$, does not affect $N$.

Our variable-density lottery model shows that it is not simply a lack of ecological realism on the part of constant relative fitness models that underlies its contrast with the classical view of density-dependent selection. Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka-Volterra equation). This precludes selection concentrated at a fragile juvenile stage facilitated by a reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.

Reproductive excesses appear in our variable-density lottery model when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting unoccupied territories survive to adulthood. Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''. The number of available slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be more sensitive to ongoing selection. In this respect, the viability selection heuristics that are common in population genetics \cite[pp. 61]{gillespie_2004} actually capture an important ecological process without making the full leap to complex age-structured models.

@@

This suggests that in systems like the wild \textit{Drosophila} example mentioned in the Introduction, there is indeed no choice but to abandon relative fitness. Our variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations for two reasons: 1) the $b$, $c$, $d$ trait scheme neatly distinguishes between different aspects of the interplay between density and selection; 2)  lottery models in general are mathematically similarity to the Wright-Fisher model, which should facilitate the analysis of genetic when $N$ is unstable.

>>>>

The constant-$N$, constant-$s$ description of selection also precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. A variety of approaches have been developed to address this in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}. Although density-dependent selection is pertinent to this longer-term issue, our focus here is the description of the time-dependent process by which selection changes allele frequencies. This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.

@@

In light of the complications arising from density-dependence, the assignment of density-independent relative fitnesses has been justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}). Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig. 9.5]{otto_2011}.

@@

Here we argue that the simplified models of density-regulated growth mentioned above are misleading in their representation of the interplay between selection and density. This ultimately derives from their failure to account for ``reproductive excess'', that is, an excess of juveniles that experience stronger selection than their adult counterparts \citep{turner1968population}. By allowing selection to be concentrated at a juvenile ``bottleneck'', reproductive excess makes it possible for the density of adults to remain constant even under strong selection. Reproductive excess featured prominently in early debates about the regulation of population density (e.g. \citealt{nicholson_1954}), and also has a long history in evolutionary theory, particularly related to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}. Additionally, reproductive excess is implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each generation involves the production of an infinite number of zygotes, of which a constant number $N$ are sampled to form the next generation of adults. Likewise in the Moran model, a juvenile is always available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.

@@

In models with detailed age structure, it is usually assumed that the density of a ``critical age group'' mediates the population's response to crowding \citep[pp. 54]{charlesworth_1994}. Reproductive excess is a special case corresponding to a critical pre-reproductive age group. A central result of the theory of density-regulated age-structured populations is that selection proceeds in the direction of increasing equilibrium density in the critical age group \citep[pp. 148]{charlesworth_1994}. This is a form of the classical $K$-selection ideas discussed above, but restricted to the critical age group (juveniles, in this case). The interdependence of pre-reproductive selection and reproductive density is thus overlooked in the process of focusing on density in the critical age group.

We re-evaluate the validity of the constant relative fitness description of selection in a novel model of density-regulated population growth that has a finite reproductive excess. Our model is inspired by the classic lottery model, which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}, and which has some similarities to the Wright-Fisher model \cite{svardal_2015}. Each type is assumed to have three traits: fecundity $b$, mortality $d$, and competitive ability $c$. In each iteration of the classic lottery model, each type produces a large number of juveniles, such that $N$ remains constant (infinite reproductive excess). Competitive ability $c$ affects the probability of winning a territory, and behaves like a pure relative fitness trait. Thus, fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). We relax the large-juvenile-number assumption of the lottery model to derive a variable-density lottery with a finite, density-dependent reproductive excess.

The properties of density-dependent selection in our model are strikingly different from the classical literature discussed above. The strong connection between crowding and selection for greater equilibrium density is broken: selection need not affect density at all. And when it does, Eq.~\eqref{eq:canonical} is almost exact even for strong selection, provided that any changes in density are driven only by selection (as opposed to large deviations from demogaphic equilibrium), and that selection occurs on only one of the traits $b$, $c$, or $d$. On the flip side, the constant relative fitness approximation fails when strong selection acts concurrently on two or more of these traits, or when the population is far from demographic equilibrium.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the difference in propagule success rate $\Delta n_2/m_2-\Delta n_1/n_1$ between an adaptive variant $2$ and a wildtype $1$ with equal frequencies. Here $b_1=1$, $d_1=0.5$ and $c_1=1$. For $b$-selection we set $b_2=b_1(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as available territories become scarce. The wildtype equilibrium density is $n_1/T\approx 0.4$, suggesting that this population would rarely experience declining strength of $c$-selection with increasing $N$.}
\end{figure}

@@

Compared to simple birth-death models \citep{kostitzin_1939} or variants of the logistic \citep{roughgarden_1979}, the Lotka-Volterra model clearly distinguishes between intra- and inter-type competitive effects. Thus, when selection acts on inter-type competitive effects, one type can displace another without having a greater equilibrium density (Fig.~\ref{fig:LVvslottery}a). This has been termed ``$\alpha$-selection'' to distinguish it from $K$-selection, which involves intra-type competitive effects and changes in equilibrium density \cite{gill_1974,joshi_2001}. Although the initial and final densities of an $\alpha$-selection sweep are the same, density nevertheless does change transiently in the Lotka-Volterra model (constant density only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012,smouse_1976}). Intuitively, for one type to exclude the other, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing $N$ to dip over a sweep (Fig.~\ref{fig:LVvslottery}a).

@@

Summarizing the properties of selection in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet, despite the differences between $b$, $c$ and $d$, in a constant environment selection that only involves one of these traits obeys the density-independent relative fitness description of selection almost exactly. When strong selection acts on more than one of $b$, $c$ and $d$ (Fig.~\ref{fig:multiple}), the constant-$s$ approximation breaks down. The $c$ and $d$ traits exemplify the two distinct directions in which density and selection can interact: selection may depend on density, and density may change in response to ongoing selection \citep{prout_1980}. The combination of both is necessary to threaten the constant-$s$ approximation. Remarkably, the $b$ trait demonstrates that the combination is not sufficient; the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.

Selection in the variable-density lottery is quite different from classical density-dependent selection (see ``Introduction'' and ``The response of density to selection; $c$-selection versus $K$-selection''). In the latter, only one life-history stage is represented, and the effects of crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka-Volterra equation). Selection in crowded populations takes broadly one of two forms: selection for greater carrying capacity ($K$-selection) or selection on competition coefficients ($\alpha$-selection). These are both ``$\delta$-like''  in the sense that selection depends on density and also causes density to change ($\delta$ is defined in Eq.~\eqref{eq:simplebirthdeath}). Strong selection is therefore sufficient for Eq.~\eqref{eq:canonical} to break down (Fig.~\ref{fig:strengthofselection}), and no distinction is made between density-regulating and density-dependent traits.

The distinctive properties of selection in the variable-density lottery arise from a reproductive excess which appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting unoccupied territories survive to adulthood. Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition. Reproductive excess produces relative contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''. The number of available slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excess, single life-stage models are biased to have total population density be more sensitive to ongoing selection. In this respect, the viability selection heuristics that are common in population genetics \cite[pp. 61]{gillespie_2004} actually capture an important ecological process without making the full leap to complex age-structured models.

Looking beyond the variable-density lottery, it is not clear which forms of crowding-induced selection are more likely to occur in nature. Even if reproductive excesses are ubquitous, strictly relative $c$-like traits could pleiotropically interact with density-regulating traits so often that $\delta$-like behavior is prevalent. For instance, in the case in the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule suggests that competitive ability is intimately linked to equilibrium resource density, and hence that $\delta$-like behavior would be prevalent. However, this conclusion is sensititive to the assumptions of well-mixed resource competition models. Spatial localization of consumable resources (e.g. for plants due to restricted movement of nutrients through soils) will tend to create territorial contests similar to the lottery model, where resource competition only occurs locally and can be sensitive to contingencies such as the timing of propagule arrival \cite{bolker_99}. In this case, resource competition is effectively subsumed into a territorial competitive ability trait akin to $c$, which would likely affect $N$ much more weakly than suggested by the $R^*$ rule (assuming no pleiotropic interactions with $b$ or $d$).

Moreover, even in well mixed populations, competition does not only involve indirect exploitation of shared resources, but also direct interference. Interference competition can dramatically alter the dynamics of resource exploitation \cite{case_1974,amarasekare_2002}, and is more likely than the exploitation of shared resource pools to involve relative contests akin to $c$-selection. For instance, sexual selection can be viewed as a form of relative interference competition between genotypes. Thus, \textit{a priori} we should not expect crowding in nature to only involve selection that is $\delta$-like. Other forms of selection like $c$-selection (that is, strictly relative traits in density-regulated populations) are also likely to be important. Note that in the classical density-dependent selection literature, interference competition is closely associated with $\alpha$-selection and the idea that selection need not affect equilibrium density \cite{gill_1974}. However, as explained above, $\alpha$-selection does transiently affect population density and can therefore be regarded as $\delta$-like.

@@

This suggests that in systems like the wild \textit{Drosophila} example mentioned in the Introduction, there is indeed no choice but to abandon relative fitness. Our variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations for two reasons: 1) the $b$, $c$, $d$ trait scheme neatly distinguishes between different aspects of the interplay between density and selection; 2) lottery models in general are mathematically similar to the Wright-Fisher model, which should facilitate the analysis of genetic drift when $N$ is unstable.
